### 2014-10-04 paper.tex
initial repo for paper

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage, times}

\begin{document}
\title{Deploying and Sharing QSAR Models using Predictive Model Markup Language}
\author{Villu Ruusmann${}^{\dagger}$ and Rajarshi Guha${}^{\ddagger}$\\
${}^{\ddagger}$National Center for Advancing Translational Sciences\\ 9800 Medical Center Drive  Rockville, MD 20850 \\ \\
${}^{\dagger}$Your address }
\date{}
\maketitle
\begin{abstract}
  A nice abstract
\end{abstract}

\section{A Plethora of Predictive Models}
\label{sec:introduction}

\section{Methods}
\label{sec:methods}

\subsection{Model Development}
\label{sec:model-development}

\subsection{Model Deployment}
\label{sec:model-deployment}

\section{Discussion}
\label{sec:discussion}

\end{document}

### 2014-10-04 termCount.py

@@ -1,0 +1,13 @@
+import urllib
+import xml.etree.ElementTree as ET
+
+u = "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi?term=%s&mindate=%d/01/01&maxdate=%d/12/31"
+term = "QSAR"
+startYear = 1980
+endYear = 2013
+for year in range(startYear, endYear+1):
+    url = u % (term.replace(" ", "+"), year, year)
+    page = urllib.urlopen(url).read()
+    doc = ET.XML(page)
+    count = doc.find("Count").text
+    print year, count

### 2014-10-04 paper.tex
Started on the introduction

@@
 \usepackage{fullpage, times}
-
+\usepackage{ctable}
 \begin{document}
@@
 
+Quantitative Structure-Activity Relationship (QSAR) models have become
+ubiquitous in computational chemistry and cheminformatics, having been
+built for a variety of endpoints including biological activities and
+physical properties. As shown in Figure \ref{fig:count-qsar} the last
+15 years has seen a significant increase in publications developing or
+using QSAR models. While there has been much discussion (REFS) on the
+prospective utility and validity of predictive QSAR models, a key
+bottleneck to their assessment is the fact that they are not easily
+obtainable. That is, the bulk of QSAR models are described in a
+publication in terms of the descriptors (sometimes made available) and
+the modeling method. While the implementation is noted, it is upon the
+reader to actually obtain the descriptor values, obtain and set up the
+modeling environment and then actually rebuild the model. In other
+words, the actual model that was developed by the authors is, usually,
+not available to be used directly. In a few cases, the model may be
+built in a pipeline tool [REF], which alleviates this problem to some
+extent. Furthermore, the model development workflow usually involves
+multiple software tools, each one of a specific version. Without
+access to the exact same version of these tools, a user cannot be
+guaranteed to exactly reproduce the model that the authors
+describe. The issue of reproducibility (of workflows and results) has
+become increasingly important [REF Greg Landrum] in the computational
+arena.
+
+A number of workers have described infrastructure to deploy predictive
+models. Some platforms such as Pipeline Pilot provide a easy to use
+method to deploy a complete pipeline as a web page. However, this
+approach does not allow one to share models with users who do not have
+access to the software. Guha [REF] described a web based
+infrastructure that allowed one to deploy serialized models developed
+in R. This architecture allowed one to distribute the model via a web
+page, allowing users to provide new input and retrieve
+predictions. But by virtue of using serialized R models, it allowed
+the user to directly work with the original model itself. An important
+downside was that the model itself did not contain explicit references
+to the descriptors that were required as input.
+
+\ctable[cap={}, caption={Growth in number of Pubmed articles that use the term QSAR.},
+label={fig:count-qsar},figure,botcap]
+{c}
+{}
+{
+  \includegraphics{img/count-qsar}
+}
+
 \section{Methods}

### 2014-10-04 termCount.py

@@ -2,7 +2,7 @@ import urllib
 import xml.etree.ElementTree as ET
 
 u = "http://eutils.ncbi.nlm.nih.gov/entrez/eutils/esearch.fcgi?term=%s&mindate=%d/01/01&maxdate=%d/12/31"
-term = "QSAR"
+term = "QSAR or QSPR"
 startYear = 1980
 endYear = 2013
 for year in range(startYear, endYear+1):

### 2014-10-04 paper.tex
Some more edis

@@
 \begin{document}
-\title{Deploying and Sharing QSAR Models using Predictive Model Markup Language}
+\title{Deploying and Sharing QSAR Models using Predictive Modeling Markup Language}
 \author{Villu Ruusmann${}^{\dagger}$ and Rajarshi Guha${}^{\ddagger}$\\
@@
 
+Bioclipse
+
+\subsection{Serializing Models}
+\label{sec:serializing-models}
+
+A key step in sharing predictive models is serialization. That is,
+converting the internal representation of the model developed in some
+environment (R, Weka, Matlab, Mathematica etc) to a disk file (usually
+a sequence of bytes). Not all deployment approaches require this step
+- usually platforms that host the deployment and development
+mechanisms (e.g., Pipeline Pilot, R using Shiny [REF]). But in general, for one to share a
+model, it must be serialized and the serialized form is exchanged.
+
+Most model development platforms support binary serialization
+schemes. For example, R can save models in the RData format [REF]
+which is binary and can only be read by the R platform (though since
+it is Open Source, one could read/write such files using custom
+code). XXX What about Matlab? XXX Java objects (say from Weka) can
+be serialized using the Java Object Protocol [REF http://docs.oracle.com/javase/7/docs/technotes/guides/serialization/]
+
+In contrast to binary serialization formats, the Predictive Modeling
+Markup Language is a plain text XML format that is designed to fully
+describe the model. Importantly, with appropriate software a PMML
+document can be used to reconstruct the model, including any data
+preprocessing steps (scaling, normalization) as well as generate input
+descriptors if appropriately specified. A number of software tools are
+capable of reading and writing PMML documents including SAS, SPSS,
+TIBCO and KNIME, whereas others such as R can produce PMML documents
+and WEKA which can consume PMML documents.
+
 \ctable[cap={}, caption={Growth in number of Pubmed articles that use the term QSAR.},
@@
 
+For this study we developed models in R 3.1 [REF] with descriptors
+generated using the CDK [REF] via the rcdk [REF] package. As exemplars
+we considered two datasets. XXX Describe the datasets.
+
+Next we developed a linear regression (or classification) and random
+forest  model  for each dataset. Using the pmml package [REF] we
+serialized the models to PMML.
+
 \subsection{Model Deployment}
@@
 
+\subsection{PMML Supports Open Science}
+\label{sec:pmml-supports-open}
+
 \end{document}

### 2014-11-20 paper.tex
The definition of a predictive solution

@@
 Bioclipse
+
+\subsection{Elements of a predictive solution}
+
+A model represents a standalone mathematical relationship. Not very easy
+to work with from the end user perspective. A model should be packaged
+as a predictive solution, which has the following components:
+\begin{itemize}
+  \item Data schema. Specifies information about input and output fields.
+  Data schema represents the "public interface" for the end user. 
+  For example, data schema could communicate that a predictive solution
+  consumes a single string-type input field that contains the SMILES 
+  representation of a chemical structure, and produces two output fields - 
+  a string-type field that contains the classification result (eg. the 
+  predicted mechanism of toxic action) and a floating point-type field 
+  that contains a class-specific regression result (eg. if the chemical
+  structure was classified as a narcotic agent, the predicted pIGC50 
+  value using the narcosis QSAR model).
+  \item Data pre- and post-processing functions. Typically, the input
+  fields that are listed in data schema cannot be directly fed to the 
+  underlying matchematical relationship "as is". Could be related to 
+  translating input from user representation to internal representation
+  (eg. converting SMILES to CDK molecular graph to CDK descriptor value)
+  or performing additional transformations on it (eg. normalization, 
+  converting from natural scale to logaritmic scale etc.).Similarly,
+  could derive secondary output fields based on the primary field. For
+  example, calculating the probability of the predicted class or standard
+  error of the predicted value.
+  \item Data (work)flows. A predictive solution may encapsulate more
+  than one model. The passage of data from one model to another happens
+  inside the predictive solution. Two most common scenarious is model 
+  aggregation, where the final prediction is calculated by applying an
+  aggregation function over a list of interim predictions (eg. consensus
+  modeling), and model chaining, where the output of earlier models
+  becomes the input of subsequent models (eg. classification followed 
+  by regression).
+\end{itemize}
+
+TODO: A figure of a chemistry-specific predictive solution.
 

### 2014-11-20 paper.tex
Dichotomy between model training and model deployment

@@
 
+\subsection{The lifecycle of a predictive solution}
+
+A modeling workflow has two clearly distinguishable lifecycle phases:
+\begin{enumerate}
+  \item Training. Happens once in the very beginning. Handled by a data
+  scientist. Typically in desktop environment. 
+  \item Deployment.Happens many times. Handled by plurality of IT 
+  roles/personnel. Typically in server-side environment, different 
+  platforms/runtimes. 
+\end{enumerate}
+
+The training is handled by a scientist. Involves analysing the
+"fundamental side" of the problem, choosing the right statistical 
+algorithms for training and validation. Also, providing the mechanistic
+interpretation of the results (white box models). The result is 
+shared with peers using modern scientific communication means. Provided
+that the result is actually useful, many people may want to deploy it
+on their premises afterwards.
+
+The deployment is an operational IT problem. Different organizations 
+may prefer different platforms, data frameworks. Therefore, the trained
+predictive solution should be maximally portable.
+
+The portability can be best acieved by relying on open and popular
+standards. To a lesser degree, if no standards have been established yet,
+could rely on open and popular tools. For example, CDK is not formally
+backed by a standard, but it is generally very easy to use and integrate
+with different chemistry products, because its source code can be obtained
+and modified easily.
+
+Overall, moving models from training phase to deployment phase is a
+major hurdle in practical data science. It is estimated that in corporate
+environments the average length of a modeling iteration may be as long
+as 16-24 months [private conversation, public ref?]. All approaches that
+can reduce this timespan and associated costs could provide significant
+business value.
+
 \subsection{Serializing Models}

### 2014-11-24 paper.tex
The uses of a published QSAR model

@@
 \usepackage{fullpage, times}
+\usepackage{hyperref}
 \usepackage{ctable}
@@
 Bioclipse
+
+\subsection{Model reproducibility vs. Model reusability}
+
+The main question: if you see a QSAR model in the literature, do you
+want to analyze its mathematical properties, or do you want to apply
+it to your dataset in order to see how it performs?
+
+One of the principles of the scientific method is that the research
+must be repeatable and reproducible. From the QSAR point of view there
+are not much technical obstacles in the way. The biggest concern is
+using proprietary algorithms and software.
+
+Model reproducibility is about marking down the procedure by which
+the model can be derived given the same input data. Theoretical methods
+do not have random error component[?]. Thus, all deviations must be
+explained by systematic errors, which could be related to choosing the
+wrong algorithm, wrong parametrization etc.
+
+Documenting the whole model training and validation process with the
+help of electronic laboratory notebooks. Should include both human and
+machine readable instructions of the process. The latter can be given 
+in the form of an executable script (ie. when executed with appropriate
+input data, produces the model and output data). Plenty of options to
+choose from (open source):
+\begin{itemize}
+  \item knitr \url{http://yihui.name/knitr/}
+  \item Sweave \url{https://www.stat.uni-muenchen.de/~leisch/Sweave/}
+  \item Sage \url{http://www.sagemath.org/}
+  \item IPython \url{http://ipython.org/}
+\end{itemize}
+
+The ONS approach is to use Wiki pages, which contain embedded R code
+and plots. For example, see
+\url{http://onschallenge.wikispaces.com/MeltingPointModel002}.
+
+Most of the time, however, the reader in interested in the \emph{final
+ending state} of the modeling workflow. That is, the mathematical 
+relationship that has been properly validated and whose applicability 
+domain has been sufficiently characterized.
+
+Essentially, the representation of a model training process boils down
+to representing the training algorithm. The most complete representation
+is in application code such as R, Python, MATLAB etc.
+The algorithm "works" by applying a finite list of well-defined 
+instructions to the initial state. After applying each instruction the
+system reaches another state. When the list of instructions has been
+exhausted the system has reached its final ending state.
+The representation of states is technically much more challenging than 
+the representation of algorithms. A state may consist of different
+parts. We are only interested in that part, which relates to the 
+mathematical relationship. A state can be preserved by taking a snapshot
+of the particular memory area. To make reading and writing more easier,
+should be formalized as a data structure.
 

### 2014-11-24 paper.tex
Capturing the states of a model training process

@@
 should be formalized as a data structure.
+
+The initial state is about the dataset. The first instructions of the
+algorithm will typically split it into training and validation subsets.
+A dataset can be represented using text-based database data formats such
+as Comma-Separated Values (CSV), Attribute-Relation File Format (ARFF)
+and others. Often becomes necessary to incorporate domain-specific
+information. For QSAR, the main need is to provide the description of
+the dependent field (a physicochemical property or biological activity)
+and independent fields (theoretical molecular descriptors). The goal is
+that if the dataset file should be "self-sustainable" if circulated
+between parties. An XML-based dataset markup language called QSAR-ML [ref?].
+
+The QsarDB approach [ref?] attempts to capture the initial state, the 
+final ending state and all states inbetween using a collection of files
+in open and standard data formats. For example, datasets are represented
+in TSV, parameter units in UCUM [ref?], bibliography in BibTeX [ref?], 
+mathematical relationships in PMML etc. QsarDB leaves the door open
+for extension information is user-defined data formats. For example, a
+QsarDB archive may capture the primary model object that corresponds to 
+the "main" run, as well as all secondary model objects that correspond to 
+validation runs.
 

### 2014-11-25 paper.tex
Three levels of model representation

@@
 
+\subsection{Representation of models}
+
+It is possible to distinguish between three levels:
+\begin{itemize}
+  \item Data structure. Tied to a specific library and/or algorithm.
+  For example, the \textit{randomForest} R library returns the trained
+  RF model as a special-purpose data structure. Moreover, the internal
+  structure of this data structure varies depending on whether the RF
+  model was trained using the "formula interface" or the "matrix interface".
+  Such data structures can be easily persisted into files using so-called
+  serialization mechanisms. The storing operation (ie. saving the in-memory
+  data structure to disk file) is called serialization, and the loading
+  operation (ie. reading the disk file data structure to memory) is called
+  deserialization. Most language runtimes provide built-in serialization
+  capabilities. For example, R language provides methods \textit{saveRDS}
+  and \textit{readRDS} that can work with binary and text-based files.
+  \item Pseudocode. A similar model can be trained using different libraries
+  and/or algorithms. For example, a RF model can also be trained using
+  the \textit{party} R library (or in-house R library). The choice of the
+  algorithm determines how the search is conducted in the feature space. 
+  However, the final ending state of a RF training process is an ensemble
+  of decision tree models.
+  The goal of the pseudocode representation is to extract and encapsulate
+  the primary structure of the model. The pseudocode representation aims 
+  to be much more informative for humans (eg. following the computation
+  path from beginning to end), but still be directly interpretable by 
+  machines. For example, a decision tree model could be represented by
+  a tree-like data structure, where non-leaf nodes specify split criteria 
+  and leaf nodes specify predicted scores.
+  \item Application code. During deployment, the model is executed with
+  with new input data. With data structure and pseudocode approaches this 
+  is accomplished by model interpretation. The alternative approach is to
+  translate either representation to application code, which would allow
+  for more optimizations and much speedier execution. 
+\end{itemize}
+
+Binary serialization mechanisms are typically much more performant and 
+compact than text serialization. The only downside, which is rather minor,
+is that binary files are hard to open and handle without special tools.
+With text files it is possible to take a peek and possibly even make minor
+modifications.
+
+The choice of serialization data format depends whether model producer 
+and model consumer will be using the same language platform or not. For
+example, the RDS data format that is default for R language is not very
+suitable when model consumer is a Java or Python application. Today,
+there are plenty multi-platform serialization data formats that offer
+bindings for most popular programming languages:
+\begin{itemize}
+  \item Protocol Buffers \url{https://code.google.com/p/protobuf/}
+  \item Apache Avro \url{http://avro.apache.org/}
+  \item XML
+  \item JSON
+  \item YAML
+\end{itemize}
+
+The first step towards pseudocode would be to have shared data structures.
+For example, from the model deployment perspective it would be desirable
+if the \textit{randomForest} and \textit{party} libraries were using common
+data structure for the representation of decision trees. In practice,
+however, it is pretty difficult to accomplish it without a third-party 
+standardizing body. The responsibilities of such third party would be 
+then to specify, develop and maintain an independent library that could
+be imported on demand. The key issue is obtaining acceptance and support
+by major tool vendors in the field. Moreover, this library of "common 
+data structures" does not have to be limited to one specific language, but
+be applicable across all platforms and languages that are used for
+everyday statistics and machine learning work.
+
+The predictive analytics community has decided to standardize around
+the Predictive Model Markup Language (PMML) [ref?], which is an XML-based
+markup language for the representation of predictive solutions.
+
 \subsection{Serializing Models}

### 2014-11-30 paper.tex
Molecular descriptor representation

@@
 
+\subsection{Representation of descriptors}
+
+A molecular descriptor is a characteristic that is obtained by applying
+a particular algorithm to a particular chemical structure representation.
+Typically classified by the complexity of structure representation:
+\begin{enumerate}
+  \item 1D. Atom counts.
+  \item 2D. Atom connectivity.
+  \item 3D. Atom locations.
+\end{enumerate}
+
+Making descriptor calculation reproducible is difficult because it is
+very tightly coupled with the preparation of chemical compounds to a suitable
+chemical structure representation. For example, consider a simplest 2D
+molecular descriptor that simply counts the number of aromcatic carbon
+atoms. The functional definition of aromaticity is still a matter of 
+ongoing debate -- classical organic chemistry interpretation (Hückel rule) 
+vs. modern theoretical/computational interpretation.
+
+By definition, when the algorithm is applied to an incorrect chemical
+structure representation, the results is an incorrect molecular descriptor
+value. Cannot reliably say if an user-specified representation is correct
+or not by merely "looking" at it. Will need to apply a rigorous verification
+procedure. For example, suppose SMILES chemical structure representation.
+There exist different variations of SMILES, some of which make aromaticy
+transparent by using different atom and bond symbols (ie. lowercase letter "c" 
+and/or carbons connected by ":" bonds). Therefore, simple analysis of SMILES
+using simple text analysis functions won't suffice, and it has to be parsed 
+and reconstructed into more sophisticated chemical graph. Even then it is
+possible to run into trouble, because the ring detection and ring aromaticity
+assignment can be performed following different algorithms.
+
+When "interfacing" QSAR models with actual end-users then it is necessary
+to specify supported input chemical structure representations. Most likely,
+actual chemical databases contain chemical compounds in data formats that
+cannot be fed \emph{directly} to the molecular descriptor calculation
+algorithm. For example, when testing with new chemical compounds then user
+input may originate from graphical drawing board (eg. JMol applet [ref?] or 
+J(S)ME Java(Script) Molecular Editor applet [ref?]) or be a systematic
+name. The desirable property of a QSAR predictive solution would be to
+recognize and automatically convert between different chemical structure
+representations.
+
+Unlike model training algorithm, the molecular descriptor calculation
+algorithm does not have intermediary reusable states. From the practical
+point of view, can be only reused by providing the complete algorithm in 
+the application code form.
+
+There have been attempts to "simplify" the systematizing and communicating 
+algorithms using the ontology approach. The earliest work is the Blue Obelisk
+Descriptor Ontology [ref?], which has been now superseded by the Chemical 
+Information (CHEMINF) Ontology [ref?]. Unfortunately, not much adoption, 
+because the vocabulary of real-life algorithms is way beyond the standardized
+vocabularies, and the lack of competing implementations. An ontology that
+has been crafted following CDK data structures and concepts is unlikely to
+be relevant and useful for other descriptor calculation software. In other
+words, if there is an one-to-one correspondence between ontology identifiers
+and CDK implementation classes, then it may be a better idea to give those
+CDK class names directly, and avoid the intermediary layers of "ontology 
+identifiers" altogether.
+
+The futility of formalizing algorithms using the ontology approach is
+well demonstrated by the "parametrization problem". The behaviour of most
+algorithms can be tweaked by adjusting a set of control variables. For 
+example, when performing the analysis of 3D molecular shapes, finding the 
+optimum between accuracy and speed (eg. mesh size). A different
+parametrization will likely result in different calculated molecular
+descriptor values. Cannot have repeatable nor less reproducible science
+when the exact parametrization is unavailable. The only way to address
+this problem would be to extend the algorithm ontology with this extra
+information, but that would limit its relevancy and usefulness even more.
+
+Another source of "variance" for molecular descriptor algorithms are 
+programming errors. It is possible that newer versions of the same software 
+could yield different results than older versions, because the algorithm
+has been slightly changed (or reparameterized) in some point in time.
+
 \section{Methods}

### 2014-12-02 paper.tex
Updated some text, added margin comments

@@
 \usepackage{ctable}
+
+%% margin notes - comment out before submission
+\usepackage{marginnote}
+\usepackage[top=2.54cm, bottom=2.54cm, outer=5cm, inner=1.25cm, heightrounded, marginparwidth=4cm, marginparsep=0.25in]{geometry}
+\renewcommand*{\marginfont}{\color{red}\small}
+
+\bibliographystyle{jcics}
+
 \begin{document}
@@
 15 years has seen a significant increase in publications developing or
-using QSAR models. While there has been much discussion (REFS) on the
+using QSAR models. While there has been much discussion
+\cite{Cherkasov:2014yq,Cramer2011Inevitable,Tropsha:2003aa} on the
 prospective utility and validity of predictive QSAR models, a key
@@
 not available to be used directly. In a few cases, the model may be
-built in a pipeline tool [REF], which alleviates this problem to some
-extent. Furthermore, the model development workflow usually involves
-multiple software tools, each one of a specific version. Without
-access to the exact same version of these tools, a user cannot be
-guaranteed to exactly reproduce the model that the authors
-describe. The issue of reproducibility (of workflows and results) has
-become increasingly important [REF Greg Landrum] in the computational
-arena.
+built in a pipeline tool \cite{KNIME:2008aa}, which alleviates this
+problem to some extent. Furthermore, the model development workflow
+usually involves multiple software tools, each one of a specific
+version. Without access to the exact same version of these tools, a
+user cannot be guaranteed to exactly reproduce the model that the
+authors describe. The issue of reproducibility (of workflows and
+results) has become increasingly important \cite{Landrum:2012hl,
+  Stodden:2013gd,Lane:2003fe} in the computational arena.
 
@@
 approach does not allow one to share models with users who do not have
-access to the software. Guha [REF] described a web based
+access to the software. Guha \cite{Guha:2007af} described a web based
 infrastructure that allowed one to deploy serialized models developed
@@
 page, allowing users to provide new input and retrieve
-predictions. But by virtue of using serialized R models, it allowed
-the user to directly work with the original model itself. An important
-downside was that the model itself did not contain explicit references
-to the descriptors that were required as input.
+predictions. While this solution allowed the deployment of models,
+they were not actually reproducible, since the serialized model did
+not actually reference the input descriptors.
 
@@
 
-The main question: if you see a QSAR model in the literature, do you
-want to analyze its mathematical properties, or do you want to apply
-it to your dataset in order to see how it performs?
+When dealing with a published QSAR model, one must choose whether to
+analyze its mathematical properties, or apply it to a dataset in order
+to see how it performs?
 
@@
 and plots. For example, see
-\url{http://onschallenge.wikispaces.com/MeltingPointModel002}.
+\url{http://onschallenge.wikispaces.com/MeltingPointModel002}. More
+recently, websites such as
+\url{https://rpubs.com/}{https://rpubs.com/} support direct web
+publishing from ones local development environment.
 
@@
 
-Essentially, the representation of a model training process boils down
+\marginnote{This paragraph reads a bit too abstract. Can it be stated
+  more concisely?}Essentially, the representation of a model training process boils down
 to representing the training algorithm. The most complete representation
@@
 
-The QsarDB approach [ref?] attempts to capture the initial state, the 
-final ending state and all states inbetween using a collection of files
-in open and standard data formats. For example, datasets are represented
-in TSV, parameter units in UCUM [ref?], bibliography in BibTeX [ref?], 
-mathematical relationships in PMML etc. QsarDB leaves the door open
-for extension information is user-defined data formats. For example, a
-QsarDB archive may capture the primary model object that corresponds to 
-the "main" run, as well as all secondary model objects that correspond to 
+The QsarDB approach \cite{Aruoja:2014bq,Ruusmann:2014uk} attempts to
+capture the initial state, the final ending state and all states
+inbetween using a collection of files in open and standard data
+formats. For example, datasets are represented in TSV, parameter units
+in UCUM [ref?], bibliography in BibTeX [ref?], mathematical
+relationships in PMML etc. QsarDB leaves the door open for extension
+information is user-defined data formats. For example, a QsarDB
+archive may capture the primary model object that corresponds to the
+"main" run, as well as all secondary model objects that correspond to
 validation runs.
@@
 
-A model represents a standalone mathematical relationship. Not very easy
-to work with from the end user perspective. A model should be packaged
-as a predictive solution, which has the following components:
+A predictive model represents a standalone mathematical
+relationship. The model itself, in terms of its mathematical form, is
+not particularly easy to work with from the end user perspective. For
+actual usage, a model should be packaged as a predictive solution,
+which has the following components:
 \begin{itemize}
@@
   and leaf nodes specify predicted scores.
-  \item Application code. During deployment, the model is executed with
+  \item Application code.\marginnote{Its not clear what this level
+      means. Are you making a distinction between 'interpretation' and
+      'compilation'? It seems data structures or pseudocode both end
+      up as application code at one point}During deployment, the model is executed with
   with new input data. With data structure and pseudocode approaches this 
@@
 
-Binary serialization mechanisms are typically much more performant and 
-compact than text serialization. The only downside, which is rather minor,
-is that binary files are hard to open and handle without special tools.
-With text files it is possible to take a peek and possibly even make minor
-modifications.
+Binary serialization mechanisms are usually more performant and 
+compact compared to text based methods. The primary disadvantage is
+that they cannot be manually inspected and require special tools to
+parse and view them.
 
@@
 a particular algorithm to a particular chemical structure representation.
-Typically classified by the complexity of structure representation:
+Typically classified by the complexity of the structure representation:
 \begin{enumerate}
-  \item 1D. Atom counts.
-  \item 2D. Atom connectivity.
-  \item 3D. Atom locations.
+  \item 1D - Atom and bond counts and other constitutional features
+  \item 2D - Atom-atom connectivity
+  \item 3D - Conformation dependent, 3D atom coordinates
 \end{enumerate}
 
-Making descriptor calculation reproducible is difficult because it is
-very tightly coupled with the preparation of chemical compounds to a suitable
-chemical structure representation. For example, consider a simplest 2D
-molecular descriptor that simply counts the number of aromcatic carbon
-atoms. The functional definition of aromaticity is still a matter of 
-ongoing debate -- classical organic chemistry interpretation (Hückel rule) 
-vs. modern theoretical/computational interpretation.
+Descriptor calculations can be difficult to reproduce. This is
+primarily driven by differences in the chemistry models implemented in
+different toolkits and chemical software platforms. An example is the
+count of aromatic carbons - which is dependent on the aromaticity
+model (of which there are several) employed by the implementation. As
+a result, while descriptors calculated using a given toolkit are
+reproducible, this is only guaranteed within that \emph{specific
+  version} of the toolkit. The same algorithm may give different
+results using another cheminformatics toolkit.
 
-By definition, when the algorithm is applied to an incorrect chemical
-structure representation, the results is an incorrect molecular descriptor
-value. Cannot reliably say if an user-specified representation is correct
-or not by merely "looking" at it. Will need to apply a rigorous verification
-procedure. For example, suppose SMILES chemical structure representation.
-There exist different variations of SMILES, some of which make aromaticy
-transparent by using different atom and bond symbols (ie. lowercase letter "c" 
-and/or carbons connected by ":" bonds). Therefore, simple analysis of SMILES
-using simple text analysis functions won't suffice, and it has to be parsed 
-and reconstructed into more sophisticated chemical graph. Even then it is
-possible to run into trouble, because the ring detection and ring aromaticity
-assignment can be performed following different algorithms.
+\marginnote{This paragraph seems to digress a little and doesn't
+  really focus on the issue at hand}By definition, when the algorithm
+is applied to an incorrect chemical structure representation, the
+results is an incorrect molecular descriptor value. Cannot reliably
+say if an user-specified representation is correct or not by merely
+"looking" at it. Will need to apply a rigorous verification
+procedure. For example, suppose SMILES chemical structure
+representation.  There exist different variations of SMILES, some of
+which make aromaticy transparent by using different atom and bond
+symbols (ie. lowercase letter "c" and/or carbons connected by ":"
+bonds). Therefore, simple analysis of SMILES using simple text
+analysis functions won't suffice, and it has to be parsed and
+reconstructed into more sophisticated chemical graph. Even then it is
+possible to run into trouble, because the ring detection and ring
+aromaticity assignment can be performed following different
+algorithms.
 
-When "interfacing" QSAR models with actual end-users then it is necessary
-to specify supported input chemical structure representations. Most likely,
-actual chemical databases contain chemical compounds in data formats that
-cannot be fed \emph{directly} to the molecular descriptor calculation
-algorithm. For example, when testing with new chemical compounds then user
-input may originate from graphical drawing board (eg. JMol applet [ref?] or 
-J(S)ME Java(Script) Molecular Editor applet [ref?]) or be a systematic
-name. The desirable property of a QSAR predictive solution would be to
-recognize and automatically convert between different chemical structure
-representations.
+\marginnote{Is it really the case that the structure representation is
+  the issue? Shouldn't any decent standardization tool transform multiple
+  input forms to a single canonical form?}Due to these issues,
+providing a ``complete'' QSAR models to an end-user requires that we
+specify the supported input chemical structure representations. Most
+likely, actual chemical databases contain chemical compounds in data
+formats that cannot be fed \emph{directly} to the molecular descriptor
+calculation algorithm. For example, when testing with new chemical
+compounds then user input may originate from graphical drawing board
+(eg. JMol applet [ref?] or J(S)ME Java(Script) Molecular Editor applet
+[ref?]) or be a systematic name. The desirable property of a QSAR
+predictive solution would be to recognize and automatically convert
+between different chemical structure representations.
 
@@
 
-There have been attempts to "simplify" the systematizing and communicating 
-algorithms using the ontology approach. The earliest work is the Blue Obelisk
-Descriptor Ontology [ref?], which has been now superseded by the Chemical 
-Information (CHEMINF) Ontology [ref?]. Unfortunately, not much adoption, 
-because the vocabulary of real-life algorithms is way beyond the standardized
-vocabularies, and the lack of competing implementations. An ontology that
-has been crafted following CDK data structures and concepts is unlikely to
-be relevant and useful for other descriptor calculation software. In other
-words, if there is an one-to-one correspondence between ontology identifiers
-and CDK implementation classes, then it may be a better idea to give those
-CDK class names directly, and avoid the intermediary layers of "ontology 
-identifiers" altogether.
+There have been attempts to "simplify" the systematizing and
+communicating algorithms using the ontology approach. The earliest
+work is the Blue Obelisk Descriptor Ontology \cite{Guha:2006ac}, which
+has been now superseded by the Chemical Information (CHEMINF) Ontology
+\cite{Hastings:2011rf}.
 
-The futility of formalizing algorithms using the ontology approach is
-well demonstrated by the "parametrization problem". The behaviour of most
-algorithms can be tweaked by adjusting a set of control variables. For 
-example, when performing the analysis of 3D molecular shapes, finding the 
-optimum between accuracy and speed (eg. mesh size). A different
-parametrization will likely result in different calculated molecular
-descriptor values. Cannot have repeatable nor less reproducible science
-when the exact parametrization is unavailable. The only way to address
-this problem would be to extend the algorithm ontology with this extra
-information, but that would limit its relevancy and usefulness even more.
+\marginnote{Probably not useful to speak negatively (at least in this
+  article). Maybe drop this paragraph or just simplify
+  it}Unfortunately, not much adoption, because the vocabulary of
+real-life algorithms is way beyond the standardized vocabularies, and
+the lack of competing implementations. An ontology that has been
+crafted following CDK data structures and concepts is unlikely to be
+relevant and useful for other descriptor calculation software. In
+other words, if there is an one-to-one correspondence between ontology
+identifiers and CDK implementation classes, then it may be a better
+idea to give those CDK class names directly, and avoid the
+intermediary layers of "ontology identifiers" altogether.
 
-Another source of "variance" for molecular descriptor algorithms are 
-programming errors. It is possible that newer versions of the same software 
-could yield different results than older versions, because the algorithm
-has been slightly changed (or reparameterized) in some point in time.
+The challenges of formalizing algorithm descriptions using ontologies
+is well demonstrated by the ``parametrization problem''.  That is, the
+behaviour of most algorithms can be tweaked by adjusting a set of
+control variables or parameters. For example, one approach to 3D
+analysis of molecular shape requires one to specify a mesh
+size. Different values represent trade-offs between speed and accuracy
+and will lead to different descriptor values. Thus a full
+specification of an algorithm must also include the details of the
+parametrization. Encoding this level of detail in an ontology may not
+be appropriate.
+
+\marginnote{This is always a possibility - not sure whether its worth
+  mentioning}Another source of "variance" for molecular descriptor
+algorithms are programming errors. It is possible that newer versions
+of the same software could yield different results than older
+versions, because the algorithm has been slightly changed (or
+reparameterized) in some point in time.
 
@@
 For this study we developed models in R 3.1 [REF] with descriptors
-generated using the CDK [REF] via the rcdk [REF] package. As exemplars
-we considered two datasets. XXX Describe the datasets.
+generated using the CDK \cite{Steinbeck:2003bh} via the rcdk
+\cite{Guha:2007aa} package. As exemplars we considered two
+datasets. XXX Describe the datasets.
 
@@
 
+\bibliography{paper}
+
 \end{document}

### 2014-12-02 paper.tex
Broad outline of the "Discussion" section

@@
 
+Putting the previous and current work of the ONS team in perspective.
+The analysis could be based on the Melting Point dataset \url{http://onswebservices.wikispaces.com/meltingpoint}, which has practical 
+applications in designing better laboratory experiments.
+
+Evolution of the Melting Point dataset \url{http://onschallenge.wikispaces.com/Open+Melting+Point+Datasets}. Over time, has been analyzed as a complete set 
+(over 10'000 chemical compounds) and chemical class-based subsets (few
+hundred to few thousand compounds).
+
+The preferred model training algorithm for the complete set has been
+Random Forest. RF model is technically quite challenging to work with,
+because it is typically several orders of magnite larger than other
+common algorithms (eg. linear regression, SVM regression).
+
+Notable accomplishments:
+\begin{itemize}
+  \item MPM001 \url{http://onschallenge.wikispaces.com/MeltingPointModel001}.
+  Analysis and visualization of the chemical space using PCA, exploratory 
+  modeling using RF vs. linear regression
+  \item MPM002 \url{http://onschallenge.wikispaces.com/MeltingPointModel002}.
+  Web service version of MPM001. PHP for frontend, R and CDK for backend.
+  \marginnote{This looks like a rather tricky setup. Intended to be a single
+  central installation that serves all interested parties?}
+  \item MPM004 \url{http://onschallenge.wikispaces.com/MeltingPointModel004}.
+  Using the high-quality subset (2706) of the complete set. The resulting RF
+  model is available for download in R native RDS data format.
+  \item MPM009 \url{http://onschallenge.wikispaces.com/MeltingPointModel009}
+  QsarDB approach. The model is exported in RDS data format instead of PMML
+  data format, but the resulting QDB archive contains CDK descriptor calculation
+  information, which makes it executable by anyone using open source QsarDB 
+  tools. Internally, the RF model data structure was converted to a generic
+  wrapper object that exposed "public API" in the form of methods getPropertyId
+  getDescriptorIds and evaluate(Map of descriptor valuess).
+  \item MPM011 \url{http://onschallenge.wikispaces.com/MeltingPointModel011}.
+  OpenBabel approach. Molecular descriptors are calculated using OpenBabel
+  software as fragment counts. The model is exported in OpenBabel data format
+  for fragment contribution models \url{http://onschallenge.wikispaces.com/file/view/mpC.txt/350881856/mpC.txt}. This model can be applied to any chemical structure
+  representation that OpenBabel supports in input. This fragment contribution 
+  table is the final model representation. The OpenBabel approach is applicable
+  only to linear models (ie. counts of atoms/bonds/fragments multiplied by their
+  contributions). Cannot do more complex model types, unless the developers
+  of the OpenBabel project decide to work on it more.
+  \item MPM012 \url{http://onschallenge.wikispaces.com/MeltingPointModel012}.
+  QSAR.io approach (ONS workbook is not completed). The RF model is represented
+  in PMML data format, and integrated with CDK descriptors over Java user-defined
+  functions (UDFs). This will make it possible to evaluate such QSAR models
+  in all Java environments, including real-time scoring, in-database scoring
+  and batch scoring.
+\end{itemize}
+
+ONS Melting Point models assume 2D chemical structure representation as input
+(eg. SMILES, InChI). It is possible to use 3D representations that are recognized
+by CDK and can be converted to the same in-memory IChemObject form (eg. Sdf).
+At later stages, ONS team began collaborating with ChemSpider and added support
+for ChemSpider Identifiers (CSIDs). When given a CSID as input, then an ONS
+webservice first resolves to an appropriate 2D representation by performing a
+query against ChemSpider database. In the future, would be equally useful to 
+support IUPAC systematic names.
+
 \subsection{PMML Supports Open Science}

### 2014-12-02 paper.tex
Categorization of predictive solutions from the end user perspective

@@
 
+\subsection{PMML export}
+
+Completeness of QSAR predictive solutions:
+\begin{enumerate}
+  \item A model, without any extras. The default option for many popular 
+  statistics applications (eg. the result of a "Save Model As ..."
+  command). Model input and output fields are underspecified, typically
+  limited only to the symbolic name of the field.
+  Very hard for end users to work with. If the training dataset is available,
+  could be even worthwhile to re-train the model (possibly using more PMML
+  capable application) and export the result with more (meta-)information so that
+  it would qualify for the second or thrid level (see below). 
+  \item A model or a model ensemble, with basic data schema, basic data pre- 
+  and post-processing. Selected statistics applications such as R and KNIME
+  provide facilities for specifying data pre- and post-processing functions
+  using PMML transformations (\url{http://www.dmg.org/v4-2-1/Transformations.html}).
+  Today, KNIME can be considered to be the most powerful statistics 
+  application for generating custom PMML documents. For example, KNIME version
+  2.9 and newer provide graphical editor for assembling workflows that produce
+  model ensembles.
+  Can be deployed by at least moderately tech-savvy end users. The biggest
+  challenge is connecting the solution to the actual data source. For example,
+  if the model was trained in R using a data.frame of CDK descriptors, then
+  the end-user must be able to calculate CDK descriptor values for the
+  "incoming" chemical compounds.
+  \item An "interaction" of models, with advanced data schema, data pre- 
+  and post-processing. The PMML document is produced programmatically by
+  custom software application. Could use a "template" approach, where the base
+  of a PMML document is hand-crafted, and model placeholder elements are
+  subsituted with fragment PMML documents that are produced by ordinary
+  statistics applications. The practical experience of one of the authors (VR)
+  suggests that the template approach is easy to set up and maintain even
+  if the overall familiarity with PMML in the organization is not that high.
+  Another customization object is reshaping the data schema of a predictive
+  solution to expected end-user needs. Moving re-defining the input fields
+  so that it becomes possible of inputting a single SMILES string instead of
+  50+ CDK descriptor values.
+\end{enumerate}
+
 \subsection{Model Development}
@@
 \label{sec:discussion}
+
+\subsection{PMML Supports Open Science}
+\label{sec:pmml-supports-open}
 
@@
 
-\subsection{PMML Supports Open Science}
-\label{sec:pmml-supports-open}
-
 \bibliography{paper}

### 2014-12-03 paper.tex
Added model pseudocode vs. model application code example

@@
 \end{itemize}
+
+Excerpt from a Random Forest model in PMML pseudocode representation:
+\begin{verbatim}
+<TreeModel splitCharacteristic="binarySplit">
+  <Node id="1">
+    <True/>
+    <Node id="2">
+      <SimplePredicate field="logP" operator="lessOrEqual" value="3.5"/>
+      <!-- More child nodes -->
+    </Node>
+    <Node id="3">
+      <SimplePredicate field="logP" operator="greaterThan" value="3.5"/>
+      <!-- More child nodes -->
+    </Node>
+  </Node>
+</TreeModel>
+\end{verbatim}
+
+The same in Java application code (using JPMML-Evaluator API) representation:
+\begin{verbatim}
+FieldValue logP = ExpressionUtil.evaluate(new FieldName("logP"), context);
+if(logP.compareToString("3.5") <= 0){ // Node 2
+  // More child if-else statements
+} else { // Node 3
+  // More child if-else statements
+}
+\end{verbatim}
 

### 2014-12-03 paper.tex
Model types in QSAR

@@
 
+\subsection{Fitness of PMML for QSAR purposes}
+Four major types of models:
+\begin{itemize}
+  \item Regression. Predicts continuous dependent value.
+  Supervised method.
+  \item Classification. Predicts categorical (unordered labels) or ordinal 
+  (ordered labels) dependent values. Two categories (binomial response) or
+  multiple categories (multinomial response). Supervised method.
+  \item Clustering. Grouping instances by similarity. Unsupervised method.
+  \item Association rules. Predicting the next instance in sequence.
+  Supervised method.
+\end{itemize}
+
+Historically, research stated around regression applications in physical organic
+chemistry such as predicting reaction rates and equilibrium constants (eg. Hammett 
+equation) using linear relationships. Later expanded into more "fuzzy" 
+application areas such as predicting biological activities (eg. Hansch equation)
+where the relationship between the dependent variable and independent variables
+is often non-linear or otherwise poorly understood (requires \textit{a posteriori} model 
+interpretation). Classification models can be often regarded as simplifications
+of regression models, where the range of a continuous response variable has been
+discretized ("binned") into small number of categories. For example, for regulatory
+purposes, it may be better to translate the predicted effect concentration based
+on safety thresholds to "not hazardous", "moderately hazardous", "highly hazardous"
+class labels. Of course, there are also many true classification problems, where
+the goal is to assign the chemical compound to one predefined chemical category.
+For example, classifying a toxicant by its structural features as "chemically inert",
+"electrophile" or "nucleophile" (without regard to its effective concentration).
+Such true classification models lead to clustering models, where the chemical space
+is dissected to subregions based on similarity measures, either structural features
+or molecular descriptor values. For example, dissecting a two-dimensional chemical
+space into four quadrants. Clustering models are the basis for read-across methods,
+where the characteristics of a new chemical compound are predicted from the 
+characteristics of its closest analogues.
+
 \subsection{Model Development}

### 2014-12-05 paper.tex
Mathematical relationships in QSAR

@@
 
+\subsection{Common mathematical relationships in QSAR}
+Three major types:
+\begin{itemize}
+  \item Standalone. The output is computed by applying a function to the input:
+  \begin{equation}
+    output = f(input)
+  \end{equation}
+  Most relationships can be efficiently represented using top-level PMML model
+  elements (15 elements as of PMML schema version 4.2). Some relationships (eg. PCA,
+  PLS) are not distinctive enough to warrant dedicated PMML model elements.
+  They are represented by combining some analogous PMML model element with appropriate 
+  data pre- or post-processing transformations. For example, a PLS model can be
+  represented by a combination of a RegressionModel element with NormContinuous 
+  data transformation elements (ie. one for each regressor).
+  \item Model aggregation. The final output is computed by applying some aggregation
+  function over a collection of interim outputs:
+  \begin{equation}
+    output = f\_aggregate(f\_1(input), f\_2(input), ..)
+  \end{equation}
+  All primary functions must be of the same type. Most common aggregation functions 
+  are the average and weighted average for regression, and the majority vote and weighted 
+  majority vote for classification and clustering. Starting from PMML schema version
+  4.0, such model aggregates can be very efficiently represented using the segmentation
+  approach (eg. a MiningModel element with a Segmentation child element).
+  Model aggregation encompasses wide range of boosting and bagging models [ref] that
+  have been gaining ground since mid 2000s and often demonstrate superior predictive
+  capabilities. For example, the Random Forest model does not warrant a dedicated PMML
+  element, because it can be completely represented as an aggregation of any number
+  (typically, in the hundreds) of decision tree models.
+  Main QSAR application is related to consensus modeling. Aggregated models can be 
+  of same type (eg. a collection of linear regression models) or different types 
+  (eg. a mix of linear regression, neural network regression and SVM regression models).
+  Such consensus models have 
+  \item Model chain. The final output is computed by applying a function to the
+  the interim output:
+  \begin{equation}
+    output = f\_2(f\_1(input))
+  \end{equation}
+  Primary functions can be of any type. Typically, a classification model chained to
+  several class-based regression models. Again, represented using the segmentation
+  approach. The output of one model may need some transformation in order to be
+  usable as input for the subsequent model.
+  Main QSAR application is mechanistic models. For example, consider the a workflow 
+  for predicting the acute aquatic toxicity of small organic molecules. First, the
+  incoming chemical compound is assigned to some class based on its predicted mechanism
+  of toxic action. Second, the the class label is used to select the corresponding
+  regression model for predicting the effective concentration (eg. pLC50, pIGC50).
+  PMML provides facilities for encapsulating model chains as a single predictive 
+  solution. From end-user perspective, much more convenient to obtain the final output
+  in single function call than to emulate it manually by evaluating multiple 
+  standalone models.
+\end{itemize}
+
 \subsection{Model Development}
